Ages of Planets and Stars Observed by Kepler

### 2024-04-19 paper/ms.tex
ms: add good new text

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}

@@

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded a detailed picture
  of how the rotation rates and lithium content of FGK stars evolve
  during the main sequence.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are collected from previous analyses of
  Kepler data; lithium is measured from new and archival HIRES
  spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is likely valid. %, and for many more for which the underlying assumptions of
%  the technique may be compromised.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report robust two-sided age measurements from both rotation and
  lithium for 14 confirmed $<$500\,Myr planets, including five
  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
  Neptune-sized planet.
  %
  In the broader Kepler target star sample,
  we find
  that the gyrochrone age distribution rises linearly from 0 to
  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
  field as stars 1\,Gyr old.
  %
  This approach offers greater sensitivity to main sequence ages than
  previous analyses of the Milky Way's star formation history, and the
  resulting age distribution bounds the relative balance of ``young''
  and ``old'' planets detectable not only by Kepler, but also by
  future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem can be solved in
various ways, and so there are many paths for discovery.  In this
manuscript, we will focus on planets discovered using transits, and
stellar ages measured based using empirical rotation and lithium-based
methods.  Rotation-based age-dating will provide the most information
for the largest number of stars.  The lithium-based technique will
provide useful information for the subset of stars younger than a few
hundred million years.

To help build intuition, it is useful to assume that the age
distribution of stars within the nearest few hundred parsecs of the
Sun (i.e.~in the Solar Neighborhood) is uniform between 0 and 10\,Gyr
\citep{Nordstrom_2004}.  We will revisit this assumption later in the
manuscript.  While the youngest 0.1\% of stars ($\lesssim$10\,million
years; Myr) are perhaps the most constraining for studies of planet
formation, they are extremely challenging targets for close-in
exoplanet detection 
\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
the early orbital, structural, and atmospheric evolution of exoplanets
are perhaps most technically feasible for the $\approx$1\% of stars
with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
younger than 1\,Gyr.  However, since stellar ages are roughly
uniformly distributed in the solar neighborhood, there is a limit on
the sample size of young planets that will be detectable in these age
regimes, particularly relative to their older breathren.

%TODO mention the RV approach?  at least the Hyades etc work...
A wide range of approaches have, over the past ten years, begun to
establish the subfield of young, close-in planet discovery.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in clusters---and
searched them for transiting planets \citep{Meibom_2013}.  Planets
discovered in this way are assumed to have the same age as the host
cluster.  The
resulting ages are precise at the $\approx$10\% level.  K2
and TESS have yielded larger numbers of young planets in clusters than
Kepler, because each of these missions has observed a larger fraction
of the sky, and therefore more stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

A limitation of the ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
This means that there are many ``young'' (0.05-1\,Gyr) stars in the
field that are omitted by cluster-only searches.
Of order 40 transiting planets are currently known to be younger than
1\,Gyr, requiring an age precision $t/\sigma_t > 3$.
%TODO cite???
% (see
%Figure~\ref{fig:rp_period_age}).
However, Kepler discovered
$\approx$4{,}000 planets; and so unless there is a major discovery
bias, we would expect a larger fraction of the Kepler
planets---perhaps 10\%, or 400 planets---to be younger than 1\,Gyr.

@@

Have these young Kepler planets already been found?  If they have, the
most likely studies to have found them would be those by
\citet{Berger_2020b_rpage}, \citet{David_2021} or
\citet{Petigura_2022}.  The \citeauthor{Berger_2020b_rpage} and
\citeauthor{Petigura_2022} studies focused on
isochronal age-dating of individual stars, which is most precise 
for stars whose luminosities and temperature separate them from the
main-sequence.  The \citeauthor{David_2021} study leveraged both isochronal
ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
relative gyrochrone age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.

Theoretical predictions for planet evolution in the Kepler sample are
most important in the context of mini-Neptunes and super-Earths, since
these are the most abundant planets.  Pertinent predictions include:
{\it a)} that mini-Neptunes should cool and contract, roughly
following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
b)} that large and close-in mini-Neptunes at the edge of the
photoevaporation desert should shrink over the atmospheric loss
timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
should decrease over $t_{\rm loss}$---the upper-envelope of the
mini-Neptune size distribution should corresponding decrease,
e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
$\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
\citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
\citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
decreasing by at least a factor of two or three, and the super-Earth
abundance increasing in tandem; {\it e)} that the
upper boundary of the super-Earth population should increase over the
same timescale as $\theta$ is increasing, since more massive
super-Earth cores should be able to retain their envelopes for longer
than less massive cores.

The details of the predictions are flexible, in that they depend on
the distributions of planetary core masses, initial atmospheric mass
fractions, and initial entropies of the atmospheres after disk
dispersal.  For instance, different sources of heating---from the star
\citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
\citep{Gupta_2019}, or even giant impacts
\citep{Biersteker_Schlichting_2019}---can produce very different
histories for the size evolution of planets at these early times
\citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
generic prediction that the ``upper envelope'' of the mini-Neptune
size distribution should shrink due to Kelvin-Helmholtz cooling is
hard to avoid; predictions however concerning the ``radius valley''
and movement of planets across it are dependent on the timescales and
prevalence of atmospheric mass loss.  In specific scenarios for
initial atmospheric accretion \citep{Lee_2022} and early atmospheric
evolution \citep{2024MNRAS.529.2716R} the radius gap might form
without needing to invoke post-disk dispersal mass loss at all.  In
such scenarios, there could be very little evolution of the planetary
size distribution between 1 and 4\,\rearth\ over the first gigayear.

Can stellar age measurements provide evidence for or against
these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While an apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
nonetheless found that when accounting for completeness, there is
$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
\citep{2024arXiv240303261V} stars than in the older Kepler sample
\citep[e.g.][]{2020AJ....159..248K}.
%TODO: more kepler citations...

Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
(e)} were explored by \citet{Berger_2020b_rpage} and the twin
\citet{David_2021} and \citet{Sandoval_2021} studies.  Both
\citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
increases with time, at a statistical significance of $\approx$2-3$\sigma$.
For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
property-matched sample of planets older than 1\,Gyr.  In
\citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
performing a Monte Carlo re-sampling procedure from the isochronal ages and
planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
effect {\it (e)} was reported by \citet{David_2021}, who found that the average
location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
$\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.

The fact that stellar ages---and especially isochronal ages---are
correlated with stellar masses and metallicities is a source of stress
for many in this subfield.  \citet{Sandoval_2021} for instance noted a
strong trend in $\theta$ as a function of stellar metallicity in
addition to the trend they noted in age, and cautioned that this could
also explain their observations.  Independently, \citet{Petigura_2018}
and \citet{Petigura_2022} have reported trends between small-planet
occurrence and both stellar metallicity and mass; these parameters
seem, at first glance, to be more important predictors of bulk changes
in the planetary population than age.  Given these known correlations,
it can be useful to consider population models in which variables such
as metallicity, mass, and age are added in succession
\citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
rigorous approach, is to jointly infer the number of planets per star
as a function of both planetary and stellar parameters, the latter
including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
most effective when the measurement precision of these different
parameters is as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
for our analysis is the fact that we can empirically calibrate
rotation-based ages using measured sequences of rotation periods
in open clusters
(\citealt{Bouma_2023}, and references therein).  This approach
ultimately ties the rotation age scale to the cluster age scale, and
is limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.
The oldest cluster for which this method has been calibrated is currently
M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.

Lithium-based age-dating includes two qualitatively distinct regimes:
depletion boundary ages derived for M-dwarfs in star clusters, and the
less precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
We summarize the origin of stellar parameters other than the ages in
Section~\ref{sec:stellarprops}, 
and describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

\section{Sample Selection}
\label{sec:selection}

This work focuses on stars surveyed by Kepler for which ages can be
inferred using either stellar rotation, lithium, or both.  These
``age-dateable stars'' are a relatively small subset of the
\nkeplerstars\ Kepler targets.  Rotation periods provide the majority
of the information, since they have been reported for
$\approx$55{,}000 Kepler stars
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  
High-resolution measurements of the \ion{Li}{1} 6708\,\AA\ doublet are
available for only a few thousand Kepler objects of interest (KOIs).
We therefore define our selection function as Kepler stars
with measured rotation periods ($\mathcal{S}$;
Section~\ref{subsec:starsel}), and Kepler planets whose host stars
have measured rotation periods ($\mathcal{P}$;
Section~\ref{subsec:planetsel}).

\subsection{Rotation periods}
\label{subsec:starsel}

>>>>

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}

@@

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded detailed
  pictures of how the rotation rates and lithium content of FGK stars
  evolve during the first few billion years of their lives.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is probably valid.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report new robust ages from both rotation and lithium for 14
  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
  super-Earths, three Earths and sub-Earths, and a Neptune-sized
  planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  The resulting age distribution bounds the relative balance of
  ``young'' and ``old'' planets detectable not only by Kepler, but
  also by future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets found using transits,
and stellar ages inferred using rotation and lithium.  Rotation
will provide the most age information for the largest number of
stars.  The lithium-based technique will provide useful information
for stars younger than a few hundred million years.

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  
The utility of this approximation is that it provides a rough estimate
for the rarity of young stars:
$\approx$1\% of stars in the nearby Galaxy have ages below 100\,Myr,
and $\approx$10\% are younger than 1\,Gyr.
Observations of exoplanets forming in disks (CITE), and of exoplanets
evolving shortly after disk dispersal (CITE CITE), are therefore
severely limited in their achievable sample sizes by the
tyranny of the Galactic star formation rate.

%While the youngest 0.1\% of stars ($\lesssim$10\,million
%years; Myr) are perhaps the most constraining for studies of planet
%formation, they are extremely challenging targets for close-in
%exoplanet detection 
%\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
%the early orbital, structural, and atmospheric evolution of exoplanets
%are perhaps most technically feasible for the 
%$\approx$1\% of stars
%with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
%younger than 1\,Gyr.  However, since stellar ages are roughly
%uniformly distributed in the solar neighborhood, there is a limit on
%the sample size of young planets that will be detectable in these age
%regimes, particularly relative to their older breathren.

%TODO mention the RV approach?  at least the Hyades etc work...

Despite the observational restrictions, the field of young close-in
planet discovery has seen remarkable success over the past ten years.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in open
clusters---and searched them for transiting planets
\citep{Meibom_2013}.  Planets discovered in this way are assumed to
have the same age as the host cluster.  The resulting ages are precise
at the $\approx$10\% level.  K2 and TESS have yielded larger numbers
of young planets in clusters than Kepler, because each of these
missions has observed a larger fraction of the sky, and therefore more
stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

A limitation of this ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
While there may be paths toward resolving this problem, this currently
means that there are many ``young'' (0.05-1\,Gyr) stars in the field
that are omitted by cluster-only searches.  Requiring ages precise to
a factor of three ($t/\sigma_t > 3$), the NASA Exolpanet Archive
currently reports $\approx$40 transiting planets younger than 1\,Gyr,
essentially all of which are in clusters.  However, Kepler discovered
$\approx$4{,}000 planets.  Unless there is a major discovery bias, we
would expect a larger fraction of the Kepler planets---perhaps 10\%,
or 400 planets---to be younger than 1\,Gyr, based on our order of
magnitude reasonsing.

@@

Have these young Kepler planets already been found?  If they have, 
it would probably have been by
\citet{Berger_2020b_rpage}, \citet{David_2021} or
\citet{Petigura_2022}.  The \citeauthor{Berger_2020b_rpage} and
\citeauthor{Petigura_2022} studies focused on
isochronal age-dating of individual stars, which is most precise 
for stars whose luminosities and temperature separate them from the
main-sequence.  The \citeauthor{David_2021} study leveraged both isochronal
ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
relative gyrochrone age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
(TODO: note that there's a lot more literature on "young" field Kepler
planets than these few sentences betray).

%      Theoretical predictions for planet evolution in the Kepler sample are
%      most important in the context of mini-Neptunes and super-Earths, since
%      these are the most abundant planets.  Pertinent predictions include:
%      {\it a)} that mini-Neptunes should cool and contract, roughly
%      following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
%      b)} that large and close-in mini-Neptunes at the edge of the
%      photoevaporation desert should shrink over the atmospheric loss
%      timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
%      abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
%      should decrease over $t_{\rm loss}$---the upper-envelope of the
%      mini-Neptune size distribution should corresponding decrease,
%      e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
%      $\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
%      \citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
%      mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
%      \citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
%      decreasing by at least a factor of two or three, and the super-Earth
%      abundance increasing in tandem; {\it e)} that the
%      upper boundary of the super-Earth population should increase over the
%      same timescale as $\theta$ is increasing, since more massive
%      super-Earth cores should be able to retain their envelopes for longer
%      than less massive cores.
%      
%      The details of the predictions are flexible, in that they depend on
%      the distributions of planetary core masses, initial atmospheric mass
%      fractions, and initial entropies of the atmospheres after disk
%      dispersal.  For instance, different sources of heating---from the star
%      \citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
%      \citep{Gupta_2019}, or even giant impacts
%      \citep{Biersteker_Schlichting_2019}---can produce very different
%      histories for the size evolution of planets at these early times
%      \citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
%      the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
%      generic prediction that the ``upper envelope'' of the mini-Neptune
%      size distribution should shrink due to Kelvin-Helmholtz cooling is
%      hard to avoid; predictions however concerning the ``radius valley''
%      and movement of planets across it are dependent on the timescales and
%      prevalence of atmospheric mass loss.  In specific scenarios for
%      initial atmospheric accretion \citep{Lee_2022} and early atmospheric
%      evolution \citep{2024MNRAS.529.2716R} the radius gap might form
%      without needing to invoke post-disk dispersal mass loss at all.  In
%      such scenarios, there could be very little evolution of the planetary
%      size distribution between 1 and 4\,\rearth\ over the first gigayear.
%      
%      Can stellar age measurements provide evidence for or against
%      these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
%      convincing---or even suggestive--- detection reported (see
%      \citealt{Petigura_2022}).  While an apparent over-abundance of large
%      very young planets ({\it (c)}) has been noted by multiple authors
%      \citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
%      exacerbated by a selection effect against sub-Neptune sized planet
%      detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
%      nonetheless found that when accounting for completeness, there is
%      $\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
%      $\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
%      \citep{2024arXiv240303261V} stars than in the older Kepler sample
%      \citep[e.g.][]{2020AJ....159..248K}.
%      %TODO: more kepler citations...
%      
%      Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
%      (e)} were explored by \citet{Berger_2020b_rpage} and the twin
%      \citet{David_2021} and \citet{Sandoval_2021} studies.  Both
%      \citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
%      increases with time, at a statistical significance of $\approx$2-3$\sigma$.
%      For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
%      sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
%      property-matched sample of planets older than 1\,Gyr.  In
%      \citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
%      performing a Monte Carlo re-sampling procedure from the isochronal ages and
%      planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
%      effect {\it (e)} was reported by \citet{David_2021}, who found that the average
%      location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
%      $\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
%      
%      The fact that stellar ages---and especially isochronal ages---are
%      correlated with stellar masses and metallicities is a source of stress
%      for many in this subfield.  \citet{Sandoval_2021} for instance noted a
%      strong trend in $\theta$ as a function of stellar metallicity in
%      addition to the trend they noted in age, and cautioned that this could
%      also explain their observations.  Independently, \citet{Petigura_2018}
%      and \citet{Petigura_2022} have reported trends between small-planet
%      occurrence and both stellar metallicity and mass; these parameters
%      seem, at first glance, to be more important predictors of bulk changes
%      in the planetary population than age.  Given these known correlations,
%      it can be useful to consider population models in which variables such
%      as metallicity, mass, and age are added in succession
%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
%      rigorous approach, is to jointly infer the number of planets per star
%      as a function of both planetary and stellar parameters, the latter
%      including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
%      most effective when the measurement precision of these different
%      parameters is as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
Today however, measurements of stellar rotation periods in open
clusters provide the most accurate empirical calibrations of
rotation-based ages (\citealt{Bouma_2023}, and references therein).
This approach ultimately ties the rotation age scale to the cluster
age scale, and is limited in precision by the intrisic scatter in the
$P_{\rm rot}$ sequences, and in accuracy by systematic uncertainties
in the evolving spin-down rates, and in the cluster ages themselves.
The oldest cluster for which this method has been calibrated is
currently M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.

Lithium ages includes two qualitatively distinct regimes: depletion
boundary ages derived for M-dwarfs in star clusters, and the less
precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our methods for selecting stellar and planetary samples in
Section~\ref{sec:selection}, and describe
the origin of our adopted stellar parameters other than ages in
Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

\section{Selecting the Stellar and Planetary Samples}
\label{sec:selection}

This work is focused on Kepler stars for which ages can be inferred
using either stellar rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods provide the majority of the information, since they have been
reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
general have only been acquired for the Kepler objects of interest
(KOIs), which comprise only a few percent of the overall stellar
sample.  In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

\subsection{Stellar Rotation Periods}
\label{subsec:rotsel}

To select stars with rotation periods, we turn to previous work.
Studies of stellar rotation across the entire Kepler field include
those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
and \citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach based on the
autocorrelation function to detect 34{,}030 rotation periods for
main-sequence Kepler targets cooler than 6500\,K, and excluded known
eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
candidates, or pulsators; they reported primary rotation periods for
24{,}124 of these stars.  \citet{Reinhold_2015} also reported
secondary periods for two thirds of the stars with primary periods,
which could be caused by differential rotation or finite spot
lifetimes.  Finally, \citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  The
rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.
